Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-27 manuscript/manuscript.tex
first SV edits

Does personal culture --an individual's attitudes, beliefs, values, and
practices--- change over the life course, or is it largely fixed once a
person reaches adulthood?

This question underlies several contemporary debates in sociology (Kiley
and Vaisey 2020; Lersch 2023; Lizardo 2017). For example, pragmatist
theories of action claim that changes in social environments cause
people to adapt their views and make new cultural meanings (Gross 2009;
Swidler 2001), while Bourdieusian practice theories argue that the
``past conditions of production'' leave a mark on people's personal
culture that lasts throughout their lives (Bourdieu 1990). Similarly,
models of social influence assume that people adapt their culture in the
face of new information, while the emphasis on cohort effects in models
of aggregate social and cultural change requires them to be open to
change while young but become fairly resistant to it as they age (Ryder
1965). Finally, life course theories posit slower but still important
changes over time as people advance through important transitions in
their lives (Bardi et al. 2009; Elder, Johnson, and Crosnoe 2003).

Because all of the above theoretical perspectives have some empirical
support, the current debate is not about whether they exist, but about
their contribution to explaining cultural differences. That said,
scholars have struggled to reach a consensus on the relative importance
of change during adulthood. Over the span of a few years, most apparent
changes in people's responses to personal survey items appear to be
transitory, with little evidence of persistent change among adults
(Kiley and Vaisey 2020; Vaisey and Kiley 2021). This suggests that
change during adulthood is not a major factor in explaining personal
cultural differences. When looked at over longer time horizons, however,
there is at least some evidence that adults make persistent cultural
changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach in
adjudicating different theoretical perspectives (Lersch 2023: p.~228).
That is, researchers have compared models of responses to cultural items
over time as consistent with either ``change'' or ``no change,'' which
essentially asks whether we observe \emph{any} intrapersonal change in
personal culture.

We argue that asking only \emph{whether} people change obscures what are
actually highly compatible empirical patterns. Being able to detect
change in a given dataset depends not only on sample populations, item
measurement, statistical methods, and power, but also on definitions of
what counts as ``change.'' These methodological decisions potentially
limit the possibility of convergence on the theoretical debates.
Moreover, asking whether there is evidence of change limits the kind of
questions that can be asked to, ``Do we see change in this measure?''
But such inquiries are ill-suited to address the core of the theoretical
debate: determining \emph{how much} the process of intrapersonal change
contributes to cultural differences.

There is enough evidence to assert that neither pre-adult socialization
nor the contemporary environment alone can fully explain cultural
differences. The question that remains is about the relative
contributions of these processes. In the current debate, for a single
construct (e.g., support for gay rights), existing approaches can only
make claims about \emph{whether} intrapersonal change is happening, not
about how much. This limitation also prevents investigating under what
conditions, across which groups, and in what domains do we see
differences in the relative importance of intrapersonal change. These
questions are about \emph{degree}, not existence, and thus require
precise measurement rather than declarations of victory for a particular
perspective.

In this paper, we intervene in this debate by moving beyond the
``tournament of models'' approach toward a quantification of the
processes that lead to differences in personal culture, and introducing
a method for quantifying the relative contributions of interpersonal
differences and intrapersonal change to the patterns we see in empirical
data. Drawing on the seven panel datasets analyzed in the debate, we use
established methods to quantify the amount of variance in responses that
is due to between-people differences at a single point during the survey
versus the amount of variance in responses that is due to linear
intrapersonal change over time. This measure gives us a sense of the
importance of change over the duration of the panel relative to the
differences that exist between people assuming they never changed.

We observe similar patterns across the datasets that yielded previously
conflicting interpretations. In all datasets we find a similar
distribution of how much explanatory power can be attributed to
intrapersonal change. But quantifying these processes also reveals that
intrapersonal change plays a substantially smaller role in explaining
the variance in peoples' personal culture than interpersonal differences
at their baselines. Whether the amount of change is large enough to be
``meaningful'' ultimately remains a theoretical question reflected in
prior expectations about particular cultural item and population. It
also crucially depends on how long people are observed in a panel.
Nevertheless, on many measures of personal culture we do not observe
enough intrapersonal change, in some cases over more than 10 years, for
it to play a substantial role in explaining the differences we observe
between adults, though notable exceptions exist.

These results help clarify and, hopefully, transcend previous debates
about whether adults change. What is more, the approach we propose can
be applied to a broader range of theoretical questions than approaches
discerning whether durable change exists in a panel. To exemplify this,
we explore differences in the explanatory power of intrapersonal change
for people with college degrees and those without. We find that change
has less explanatory power for people with college degrees than those
without. This may imply that the college experience solidifies political
dispositions rather than fosters an openness and adaptability to new
information and influences later in life. This offers a novel, empirical
basis to theorize about the role of college in shaping personal culture.
We believe that this approach has the potential to advance the
theoretical debate on the role of environments, social identities, and
roles in facilitating personal cultural change and stability.

@@

Third, we expect the variance explained by each process
-\/--interpersonal differences and intrapersonal change-\/-- to depend
on specific attributes of the question and panel. First, the higher the
response resolution for a given question and the larger the sample, the
greater the variance explained by intrapersonal change should be, as
smaller changes in intrapersonal change will be easier to detect.
Second, the LCAM and various theoretical perspectives that underlie it
posit that social transitions drive changes in personal culture. Since
the probability that any person experiences such a transition goes up
the longer we observe them, we expect panel duration to be positively
related to the proportion of variation explained by intrapersonal
change.

@@

We use data from 7 nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items that capture attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-cultural sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Appendix A documents the list of all variables used in
the upcoming analyses.

@@

As noted above, the substantive importance of intrapersonal change for
understanding cultural differences depends on a range of factors,
including how long the panel runs, whether assumptions about linear
change hold, and whether the panel is capturing a distinctly turbulent
period or a distinctly stable period for the item under question.
However, if we assume that the period under observation is ``typical''
for a question ---not a time of extremely heightened (or lowered)
sensitivity or change-\/-- then it does not seem realistic for this
process of intrapersonal change to explain a large share of
interpersonal differences for many of the questions we observe.

@@

@@

\textit{Notes:} The figure shows $\omega$ and 1 - $\omega$ as the proportion of systematic variance attributable to intrapersonal change and interpersonal differences, respectively. See Appendix A for the full set of item values. 
\end{figure}

@@

\textit{Notes:} The model estimates the proportion of systematic variance attributable to intrapersonal change. It includes the log of the number of participants, the number of waves the variable was asked (ceiled at t = 10 for each item), the date of the first wave (measured as the year minus 1968 divided by 10), number of waves the question was asked averaged across participants, duration in years per item averaged across participants, and panel and topic fixed effects. Coefficients are estimated from suppressed intercept model based on predictions at participants' wave mid-point. Survey indicators and item topics not shown. See Appendix B for the full set of coefficients.
\end{figure}

>>>>

Does personal culture---an individual's attitudes, beliefs, values, and
practices---change over the life course, or is it largely fixed by
adulthood? This question underlies an important contemporary debate in
sociology (Kiley and Vaisey 2020; Lersch 2023; Lizardo 2017) and has
deep roots in seemingly contradictory theoretical perspectives. For
example, pragmatist theories of action claim that changes in social
environments cause people to adapt their views and make new cultural
meanings (Gross 2009; Swidler 2001), while Bourdieusian practice
theories argue that the ``past conditions of production'' leave a mark
on people's personal culture that lasts throughout their lives (Bourdieu
1990). Models of social influence assume that people adapt their culture
in the face of new information (Goldberg and Stein 2018), while the
emphasis on cohort effects in models of aggregate social and cultural
change requires them to be open to change while young but become fairly
resistant to it as they age (Ryder 1965). Finally, life course theories
posit important changes over time as people advance through important
transitions in their lives (Bardi et al. 2009; Elder, Johnson, and
Crosnoe 2003).

Because all of these theoretical perspectives have some empirical
support, the current debate is not about whether their posited processes
exist at all, but about their relative contribution to explaining the
cultural differences we see in the world. That said, scholars have
struggled to reach a consensus on the importance of change during
adulthood. Over the span of a few years, most apparent changes in
people's responses to personal survey items appear to be transitory,
with little evidence of persistent change among adults (Kiley and Vaisey
2020; Vaisey and Kiley 2021). This suggests that change during adulthood
is not a major factor in explaining personal cultural differences. When
we consider a longer time horizon, however, there is evidence that
adults make at least some persistent changes (Lersch 2023).

These seemingly inconsistent findings are partly due to the fact that
researchers have taken a ``tournament of models'' approach to
adjudicating different theoretical perspectives (Lersch 2023: p.~228).
That is, researchers have compared the fit of different statistical
models to each item to determine whether it is more consistent with ``no
change'' or ``any change.'' The size of these piles then serves as the
primary evidence for the truth of the theory (Kiley and Vaisey 2020;
Lersch 2023; Vaisey and Kiley 2021).

We argue, however, that asking \emph{whether} people change on an item
may cause more problems than it solves. Being able to detect \emph{any}
change on an item depends not only on sample populations, measurement,
statistical methods, and power, but also on definitions of what counts
as ``change.'' These differences can limit the possibility of agreement
on which ``pile'' a variable belongs in. Moreover, asking whether there
is evidence of change reduces to, ``Is there \emph{any} evidence of
change in this item?'' But such binary inquiries are ill-suited to
address the core of the theoretical debate: determining \emph{how much}
the process of intrapersonal change contributes to cultural differences.

There is enough evidence to assert that neither pre-adult socialization
nor the contemporary social context alone can explain the full range of
cultural differences. The question that remains is about the
\emph{relative} contributions of these processes. In the current debate,
for a single construct (e.g., support for gay rights), existing
approaches can only make claims about \emph{whether} intrapersonal
change is happening, not about how much. This also prevents
investigating under what conditions, across which groups, and in what
domains do we see differences in the relative importance of
intrapersonal change and pre-existing interpersonal differences. These
questions are about \emph{degree}, not existence; taking this debate
forward will require precise measurement rather than declarations of
victory for a particular perspective.

In this paper, we move beyond the ``tournament of models'' approach and
introduce a method for quantifying the relative contributions of
interpersonal differences and intrapersonal change for a single item.
Drawing on seven panel data sets from five countries, we quantify the
amount of variance in survey responses that is due to between-person
differences at a single point during the survey versus the amount of
variance in responses that is due to systematic intrapersonal change
over the study period. This measure quantifies the amount of change over
the duration of the panel relative to the stable differences that exist
between people.

Using this measure, we observe similar patterns across the datasets that
have previously yielded conflicting interpretations. In all
datasets---despite their varied social contexts and duration---we find a
similar distribution of intrapersonal change. In general, we find that
intrapersonal change accounts for a substantially smaller amount of
variance in personal culture survey item than stable interpersonal
differences. Of course, whether the amount of intrapersonal change in an
item is large enough to be ``meaningful'' is a substantive and
theoretical question. Nevertheless, on many measures of personal culture
(some observed for over a decade), we do not observe enough
intrapersonal change for it to play a substantial role in explaining the
differences we observe between adults. Notable, interesting, and
important exceptions exist, of course, but they are departures from the
overall pattern.

Our method can also help address substantive questions about personal
change. To illustrate this, we investigate differences in the relative
importance of intrapersonal change on political items for people with
and without a college degree. We find that change is less pronounced for
college-educated respondents, suggesting that college solidifies
political dispositions rather than fostering an openness to new
information. This offers a novel empirical basis to theorize about the
role of college in shaping personal culture . In summary, we believe
that this approach has the potential to advance the theoretical debate
on the role of environments, social identities, and roles in
facilitating personal cultural change and stability.

Taken together, to goal of this paper is to advance---and, hopefully,
transcend---debates about \emph{whether} adults change. They do change,
at least a little, on most things. Our method can help quantify exactly
how much.

@@

Third, we expect the variance explained by each process ---interpersonal
differences and intrapersonal change--- to depend on specific attributes
of the question and panel. First, the higher the response resolution for
a given question and the larger the sample, the greater the variance
explained by intrapersonal change should be, as smaller changes in
intrapersonal change will be easier to detect. Second, the LCAM and
various theoretical perspectives that underlie it posit that social
transitions drive changes in personal culture. Since the probability
that any person experiences such a transition goes up the longer we
observe them, we expect panel duration to be positively related to the
proportion of variation explained by intrapersonal change.

@@

We use data from 7 nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items that capture attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-cultural sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

@@

As noted above, the substantive importance of intrapersonal change for
understanding cultural differences depends on a range of factors,
including how long the panel runs, whether assumptions about linear
change hold, and whether the panel is capturing a distinctly turbulent
period or a distinctly stable period for the item under question.
However, if we assume that the period under observation is ``typical''
for a question ---not a time of extremely heightened (or lowered)
sensitivity or change--- then it does not seem realistic for this
process of intrapersonal change to explain a large share of
interpersonal differences for many of the questions we observe.

@@

\leavevmode\vadjust pre{\hypertarget{ref-goldberg2018}{}}%
Goldberg, Amir, and Sarah K. Stein. 2018. {``Beyond {Social}
{Contagion}: {Associative} {Diffusion} and the {Emergence} of {Cultural}
{Variation}.''} \emph{American Sociological Review} 83(5):897--932. doi:
\href{https://doi.org/10.1177/0003122418797576}{10.1177/0003122418797576}.

@@

\textit{Notes:} The figure shows $\omega$ and 1 - $\omega$ as the proportion of systematic variance attributable to intrapersonal change and interpersonal differences, respectively. See Supplemental Materials A for the full set of item values. 
\end{figure}

@@

\textit{Notes:} The model estimates the proportion of systematic variance attributable to intrapersonal change. It includes the log of the number of participants, the number of waves the variable was asked (ceiled at t = 10 for each item), the date of the first wave (measured as the year minus 1968 divided by 10), number of waves the question was asked averaged across participants, duration in years per item averaged across participants, and panel and topic fixed effects. Coefficients are estimated from suppressed intercept model based on predictions at participants' wave mid-point. Survey indicators and item topics not shown. See Supplemental Materials B for the full set of coefficients.
\end{figure}
